---
title: Random matrix generation
author: ''
date: '2020-03-17'
slug: random-matrix-generation
categories:
  - Research
tags:
  - R Code
---

For better or worse, having lots of time inside during the [COVID-19 Pandemic](https://www.cdc.gov/coronavirus/2019-ncov/index.html) has led me to tackle some of my outstanding R coding problems.
Here's one possible solution I've come up with for generating random community abundance matrices following one of the suggested approaches in [Ulrich and Gotelli 2010](https://esajournals.onlinelibrary.wiley.com/doi/abs/10.1890/09-2157.1).

# Implementing the IT random matrix filling / creation algorithm from Ulrich and Gotelli 2010, Ecology.

To quote from the text:

> IT assigns individuals randomly to matrix cells with probabilities proportional to observed row and column abundance totals until, for each row and column, total abundances are reached.

## Approach

To sample with "probabilities proportional to observed row and column abundance totals", I used the `chisq.test` function. 
The expected values in a matrix (or contingency table) are a function of the row and column abundances.
I believe this approach is supported by Ulrich and Gotelli's comparison of the results they found using the IT algorithm compared to $\chi^2$ analysis for 200 random matrices (p. 3395).

Using these **expected** values, we can generate matrices comprised of all zero values *except* a single 1 value for a random matrix element. 
The element that is assigned a value of 1 will be decided based on a random draw from the matrix indices that is weighted by the *expected* values in the original matrix.
Next, by adding a series of these random matrices we can make a new matrix with the same **matrix sum** as the original matrix.
Finally, using an `if` statement, we can do this while assuring that the row and column abundances remain the same as the original matrix.

## Working through an example 

Below I present two examples of using this approach. The first comes from the help file for the `chisq.test` function, with some small modifications.

### `chisq.test` example

First thing to do is to use `chisq.test` to get the **expected** counts for any matrix. 
These expected counts are a function of the row and column sums:

$$
\text{expected count} = \frac{\text{row sum} \times \text{column sum}}{\text{matrix sum}}
$$

```{r}
M <- as.table(rbind(c(762, 327, 468), c(484, 239, 477)))
Xsq <- chisq.test(M)  # Prints test summary
# Xsq$observed   # observed counts (same as M)
Xsq$expected   # expected counts under the null
```

Next, we use a `while` loop to make a new matrix with a matrix sum equal to that of the original matrix.
Within the `while` loop, new matrices are created with all zeros *except* a randomly placed value of 1. 
We're adding an `if` statement to "throw out" new zero matrices that result in row or column sums that are greater than those for the original matrix.

```{r}
# Make a new matrix of all zeros
new_mat <- M * 0

while(sum(new_mat) < sum(M)){
  zero_mat <- M * 0
  zero_mat[sample(1:length(Xsq$expected), size = 1, prob = Xsq$expected)] <- 1
  new_mat_temp <- new_mat + zero_mat
  if( !any((rowSums(new_mat_temp) - rowSums(Xsq$expected)) > 0 ) &
      !any((colSums(new_mat_temp) - colSums(Xsq$expected)) > 0 )) {
    new_mat <- new_mat_temp
  }
}
new_mat

```

Finally, let's do just a few checks to make sure the matrix sums are the same and the row and column sums are the same.

```{r}
sum(new_mat) == sum(M)
all(rowSums(new_mat) == rowSums(Xsq$expected))
all(colSums(new_mat) == colSums(Xsq$expected))

```


### A larger random matrix

Here I'm applying the same setup as above on a 10 x 10 matrix, with shuffled values between 1 and 100.

```{r}
rand_matr <- matrix(sample(1:100), nrow = 10, ncol = 10)
rand_matr_exp <- chisq.test(rand_matr)$expected 

# Make a new matrix of all zeros
new_mat <- rand_matr * 0

while(sum(new_mat) < sum(rand_matr)){
  zero_mat <- rand_matr * 0
  zero_mat[sample(1:length(rand_matr_exp), size = 1, prob = rand_matr_exp)] <- 1
  new_mat_temp <- new_mat + zero_mat
  if( !any((rowSums(new_mat_temp) - rowSums(rand_matr_exp)) > 0 ) &
      !any((colSums(new_mat_temp) - colSums(rand_matr_exp)) > 0 )) {
    new_mat <- new_mat_temp
  }
}
new_mat

sum(new_mat) == sum(rand_matr)
all(rowSums(new_mat) == rowSums(rand_matr_exp))
all(colSums(new_mat) == colSums(rand_matr_exp))


```

## Making a function for this

Let's make this into a function

```{r}
it_randomize <- function(matr){
  
  # perform chisq test to get expected values
  matr_exp <- chisq.test(matr)$expected 
  
  # make a new matrix of all zeros
  new_mat <- matr * 0
  
  # use a while loop to add matrices with a single, randomly
  # placed value of 1 until
  # the new_mat sum is the same as the original matrix
  while(sum(new_mat) < sum(matr)){
    # make a random matrix
    # this is done by randomly choosing a matrix index value, weighted
    # by the expected matrix values at each index
    zero_mat <- matr * 0
    zero_mat[sample(1:length(matr_exp), size = 1, prob = matr_exp)] <- 1
    
    # add the random matrix to the new_mat temporalily
    new_mat_temp <- new_mat + zero_mat
    
    # keep new_mat_temp if it doesn't result in over filling the 
    # row or column sums
    if( !any((rowSums(new_mat_temp) - rowSums(matr_exp)) > 0 ) &
        !any((colSums(new_mat_temp) - colSums(matr_exp)) > 0 )) {
      new_mat <- new_mat_temp
    }
  }
  
  return(new_mat)
}

```

Apply the new function

```{r}
new_mat <- it_randomize(M)
new_mat
sum(new_mat) == sum(M)
all(rowSums(new_mat) == rowSums(Xsq$expected))
all(colSums(new_mat) == colSums(Xsq$expected))

```


```{r}
new_mat <- it_randomize(rand_matr)
new_mat
sum(new_mat) == sum(rand_matr)
all(rowSums(new_mat) == rowSums(rand_matr_exp))
all(colSums(new_mat) == colSums(rand_matr_exp))

```

